---
title: "S&P 500: Great Depression vs. 2008 Financial Crisis"
author: "William Chiu"
date: "5/17/2020"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

The stock market recovery from the Great Depression (of 1929) was very slow compared to the recovery from the Financial Crisis (of 2008). The document compares the daily performance of `$10,000` invested one month prior to the beginning of each recession.

A portfolio of `$10,000` at the beginning of the Great Depression never recovers even after 12.5 years. In stark contrast, a portfolio of `$10,000` recovers approximately 5.3 years after the Financial Crisis.

I thank the author of this [R tutorial](https://www.codingfinance.com/post/2018-03-27-download-price/). I also want to thank the author of this [other tutorial](https://blog.exploratory.io/introducing-time-series-analysis-with-dplyr-60683587cf8a). And  [this tutorial](https://stackoverflow.com/questions/23518605/add-an-index-numeric-id-column-to-large-data-frame/23518737). One of the best things about using R is the ease of copying and pasting other people's code.

Advanced users may adjust the ticker symbol and initial investment dates in the chunk below and re-run the analysis.
```{r}
### document parameters

ticker <- "^GSPC"

GDstartdate <- '1929-07-01'
FCstartdate <- '2007-11-01'
```


# Data

Extract historical data for the S&P 500 from Yahoo Finance.

```{r, message=FALSE}
library(tidyverse)
library(tidyquant)

GSPC <- tq_get(ticker, from = '1927-01-01',
           to = "2020-05-17",
               get = "stock.prices")

head(GSPC)
```

# Recession Dates

According to the [NBER](https://www.nber.org/cycles.html) the Great Depression began on `August 1929` while the 2008 Financial Crisis began on `December 2007`. In order to simulate the performance of a portfolio that began immediately prior to the beginning of the recession, the document assumes `$10,000` was invested in the S&P 500 on the following dates:

* 1929-07-01
* 2007-11-01

```{r}
GSPC %>% 
  filter(date >= GDstartdate) %>% 
  head()

GSPC %>% 
  filter (date >= FCstartdate) %>% 
  head()
```

# Splitting the Data

Split the data sets between Great Depression and Financial Crisis. Then add new variable for portfolio age (i.e., the number of days since the initial `$10,000` investment). Use calendar days rather than trading days.

```{r}
GD <- GSPC %>% 
  filter(date >= GDstartdate) %>% 
  mutate(PortfolioAge = interval(first(date), date) / days(1))

FC <- GSPC %>% 
  filter (date >= FCstartdate) %>% 
  mutate(PortfolioAge = interval(first(date), date) / days(1))

head(GD)

head(FC)
```

# Calculate the cumulative returns

```{r}
GD2 <- GD %>% 
  mutate(CumReturn1929 = adjusted / first(adjusted)) %>% 
  select(PortfolioAge, date, adjusted, CumReturn1929)

FC2 <- FC %>% 
  mutate(CumReturn2008 = adjusted / first(adjusted)) %>% 
  select(PortfolioAge, date, adjusted, CumReturn2008)

head(GD2)
head(FC2)

```

# Merge the returns by Portfolio Age

```{r}
GD_FC <- GD2 %>% 
  inner_join(FC2, by="PortfolioAge") %>% 
  select(PortfolioAge, CumReturn1929, CumReturn2008)

head(GD_FC)
```

We're almost done. We need to multiply the cumulative returns by `$10,000` to get the value of the portfolio in dollar terms.

```{r}
GD_FC2 <- GD_FC %>% 
  mutate(GreatDepression1929 = CumReturn1929*10000,
          FinancialCrisis2008 = CumReturn2008*10000)

head(GD_FC2)
```

# Convert columns to rows

This part is a little tricky. The columns `GreatDepression1929` and `FinancialCrisis2008` should be converted into rows with a new column (i.e., `Type`) to distinguish between the 1929 and 2008 portfolios. I perform this step to make plotting easier.

```{r}
GD_FC2_gather <- GD_FC2 %>% 
  select(PortfolioAge, GreatDepression1929, FinancialCrisis2008) %>% 
  gather("Type", "PortfolioValue", -PortfolioAge)

head(GD_FC2_gather)

tail(GD_FC2_gather)
```

# Plot

```{r}
ggplot(GD_FC2_gather, aes(x=PortfolioAge, y=PortfolioValue)) +
  geom_line(aes(color=Type)) +
  theme_classic() +
  geom_hline(yintercept=10000, linetype="dashed", color = "black") +
  scale_y_continuous(breaks=seq(0,25000,2500)) + 
  scale_x_continuous(name="Portfolio Age (in days)",breaks=seq(0,365*13,365)) +
  scale_color_manual(values=c('blue','orange')) +
  ggtitle("Investing $10,000 in the S&P 500")

```

# Conclusion

Had you invested `$10,000` on `1929-07-01`, your portfolio would have been worth about `$3,000` after 12.5 years. In stark contrast, if you invested `$10,000` on `2007-11-01`, your portfolio would have been worth about `$19,000` after 12.5 years.
